Syntax and Semantics of Cedille

### 2020-12-22 docs/semantics/paper.tex
docs: semantics: changes to several typing rules
- use \varphi instead of \phi
- use \rho with guide
- remove occurences of types being checked against kinds
- only one version of chi

    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t \simeq t \}}{\Gamma\vdash \textit{FV}(t\ t')\subseteq \textit{dom}(\Gamma)}  \\ \\    
    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
    \infer{\Gamma\vdash \rho\ t'\ \mhyph\ t \Leftrightarrow [t_2/x]T}
          {\Gamma\vdash t' \tpsynth \{ t_1 \simeq t_2 \} & \Gamma \vdash t \Leftrightarrow [t_1/x]T} \\ \\
    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpcheck T'}
          {\Gamma\vdash T\tpcheck \star & \Gamma\vdash t \tpcheck T & T \cong T'} &
    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpsynth T}
          {\Gamma\vdash T\tpcheck \star & \Gamma\vdash t \tpsynth T' & T \cong T'} \\ \\
    \infer{\Gamma\vdash \phi\ t\ \mhyph\ t'\{t''\} \Leftrightarrow T}
          {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}  & \
  \end{array}
  \]
\caption{Rules for checking a term against a well-kinded type ($\Gamma \vdash t \tpcheck T$)
           and synthesizing a type for a term ($\Gamma \vdash t \tpsynth T$)}
\label{fig:tp}
\end{figure}

\begin{figure}
  \[
  \begin{array}{lllllll}
    |x| & = & x &\ &
    |\abs{\lambda}{x}{t'}{t}| & = & \absu{\lambda}{x}{|t|} \\
    |t\ t'| & = & |t|\ |t'| &\ &
    |t\cdot T| & = & |t| \\
    |\abs{\Lambda}{x}{t'}{t}| & = & |t| &\ &
    |t\ \mhyph t'| & = & |t| \\
    |[t , t']| & = & |t| &\ &
    |t.1| & = & |t| \\
    |t.2| & = & |t| &\ &
    |\beta\{t\}| & = & |t|\\
    |\delta\ \mhyph\ t| & = & \absu{\lambda}{x}{x}&\ &
    |\rho\ t\ \mhyph\ t'| & = & |t'| \\
    |\phi\ t\ \mhyph\ t'\ \{t''\}| & = & |t''| &\ &
    |\chi\ T\ \mhyph\ t'| & = & |t'|
  \end{array}
  \]
  \caption{Erasure for annotated terms}
  \label{fig:eraser}
\end{figure}

@@

We have modified the rules for equality types $\{ t \simeq t' \}$ so that we require
nothing of $t$ and $t'$ except that the set $\textit{dom}(\Gamma)$ of variables
declared by $\Gamma$ includes their free variables $\textit{FV}(t\ t')$.  Further modifications
over the version of CDLE in~\cite{stump18} are:
\begin{itemize}
\item To prove $\{ t \simeq t\}$, one now writes $\beta\{t'\}$,
  with the critical idea that $|\beta\{t'\}|$ erases to $|t'|$.  We call this the \textbf{Kleene trick},
  because it goes back to Kleene's numeric
  realizability~\cite{Kle65_Classical-Extensions-of-Intuitionistic-Mathematics},
  which accepts any number $n$ as a realizer 
  of a true equation.  Here, we accept any closed term $t$ as a realizer
  of $\{ t \simeq t\}$.  This means that in Cedille, any such term -- even otherwise untypable
  terms, non-normalizing terms, etc. -- have type $\{ t \simeq t\}$ for any term $t$.
\item The $\rho$ construct allows one to rewrite occurrences of $t_1$ to $t_2$ in the synthesized or
  checked type, where $t_1$ and $t_2$ are provably equal.  In the Cedille implementation, we rewrite
  all matching occurrences.  This may be compared to \verb|rewrite| in Agda, except that it may be
  applied anywhere, not just as part of pattern matching~\cite{agda}.
\item We adopt a strong form of Nuprl's \textbf{direct computation rules}~\cite{constable+86}:
  If we have a term $t'$ of type $T$ and a proof $t$ that $\{ t' \simeq t''\}$, then we may conclude that
  $t''$ has type $T$ by writing the annotated term $\phi\ t\ \mhyph\ t'\{t''\}$, which
  erases to $t''$.
\item Where the previous version of CDLE uses $\beta$-equivalence for (erased) terms, we here adopt $\beta\eta$-equivalence.  This
  allows us to observe in many cases that retyping functions are actually $\beta\eta$-equivalent to $\absu{\lambda}{x}{x}$.
  While $\beta\eta$-equivalence takes more work to incorporate into intrinsic 
  type theory~\cite{geuvers92},
  it raises no difficulties for our extrinsic one.
\item In this version, we add an explicit axiom $\delta$ saying that Church-encoded boolean \emph{true} is different from
  \emph{false}.
  In the implementation of Cedille, this is generalized to the rule where the
  proof \(t\) synthesizes an equation \(\{t_1 \simeq t_2\}\) in which \(|t_1|\)
  and \(|t_2|\) are separable using the \emph{B\"ohm-out algorithm}~\cite{BDPR79_Bohm-Algorithm}.

@@

Defining \verb|Top| to be $\{\absu{\lambda}{x}{x} \simeq
\absu{\lambda}{x}{x}\}$, we may assign \verb|Top| to any closed term \verb|t|,
including non-normalizing ones. In our annotated syntax, we write
\texttt{\(\beta\)\{t\}}. Even without this, the presence of $\delta$ in
combination with $\phi$ allows us to type non-normalizing terms assuming an
erased argument $x$ of type $\{ \textit{tt} \simeq \textit{ff} \}$ for
Church-encoded booleans \textit{tt} and \textit{ff}. For example, $\delta\ \mhyph x$
has type \(\{\absu{\lambda}{x}{x} \simeq \absu{\lambda}{x}{x\ x}\}\), and with
$\phi$ we can use this to type $\Omega$ by changing the typed term
\(\texttt{id}\ \cdot\texttt{True}\ \texttt{id}\), where \verb|True| is
\(\abs{\forall}{X}{\star}{X \to X}\). But failure of normalization does not
impinge on Theorem~\ref{thm:consis}. Extensional Martin-L\"of type theory (MLTT)
is also non-normalizing, for a very similar reason, but this fact does not contradict
its logical soundness~\cite{dybjer16}. In CDLE, the guarantees one gets about
the behavior of terms are expressed almost entirely in their types. If the types
are weak, then not much is guaranteed; but stronger types can guarantee
properties like normalization, as demonstrated by the following theorem:

@@

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpcheck T'}
          {\Gamma\vdash T\tpcheck \star & \Gamma\vdash t \tpcheck T & T \cong T'} 
\]
Using the IH for the first premise and the assumption that $\interp{T'}_{\sigma,\rho}$ is in $\mathcal{R}$ and hence defined,
we can apply the IH to the third premise to get $\interp{T}_{\sigma,\rho} = \interp{T'}_{\sigma,\rho}$.  Using this and the IH for second premise,
we get the desired conclusion, using also the definition of erasure.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpsynth T}
          {\Gamma\vdash T\tpcheck \star & \Gamma\vdash t \tpsynth T' & T \cong T'} 
\]
By the IH applied to the second premise, we have $[\sigma|t|]_{\cbe}\in\interp{T'}_{\sigma,\rho}\in\mathcal{R}$.
Using definedness of $\interp{T'}_{\sigma,\rho}$ and the IH applied to the first premise, we can apply the
IH to the third premise to get $\interp{T}_{\sigma,\rho} = \interp{T'}_{\sigma,\rho}$, from which the desired
conclusion follows by definition of erasure.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \phi\ t\ \mhyph\ t'\{t''\} \Leftrightarrow T}
          {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}  
\]
By the IH for the first premise, $\sigma|t'| =_{\cbe} \sigma|t''|$.  By the IH for the second
premise, $[\sigma|t'|]_{\cbe}\in\interp{T}_{\sigma,\rho}$.  This suffices for the desired conclusion,
using also the definition of erasure ($|\phi\ t\ \mhyph\ t'\{t''\}| = |t''|$).

>>>>

    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t \simeq t \}}{\Gamma\vdash \textit{FV}(t')\subseteq \textit{dom}(\Gamma)}  \\ \\    
    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
    \infer{\Gamma \vdash \rho\ t'\ @x.T\ \mhyph\ t \tpsynth [t_2/x]T}
          {
          \begin{array}{cc}
            \Gamma \vdash t' \tpsynth \{t_1 \simeq t_2\}
            & \quad \Gamma \vdash [t_2/x]T \tpsynth \star
            \\ \Gamma \vdash t \tpsynth T'
            & T' \cong [t_1/x]T
          \end{array}
          } \\ \\
    \infer{\Gamma \vdash \chi\ T\ \mhyph t\ \tpsynth T}
          {\Gamma \vdash T \tpsynth \star & \Gamma \vdash t \tpcheck T} &
    \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\{t''\} \Leftrightarrow T}
          {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}
  \end{array}
  \]
\caption{Rules for checking a term against a well-kinded type ($\Gamma \vdash t \tpcheck T$)
           and synthesizing a type for a term ($\Gamma \vdash t \tpsynth T$)}
\label{fig:tp}
\end{figure}

\begin{figure}
  \[
  \begin{array}{lllllll}
    |x| & = & x &\ &
    |\abs{\lambda}{x}{t'}{t}| & = & \absu{\lambda}{x}{|t|} \\
    |t\ t'| & = & |t|\ |t'| &\ &
    |t\cdot T| & = & |t| \\
    |\abs{\Lambda}{x}{t'}{t}| & = & |t| &\ &
    |t\ \mhyph t'| & = & |t| \\
    |[t , t']| & = & |t| &\ &
    |t.1| & = & |t| \\
    |t.2| & = & |t| &\ &
    |\beta\{t\}| & = & |t|\\
    |\delta\ \mhyph\ t| & = & \absu{\lambda}{x}{x}&\ &
    |\rho\ t\ @x.T\ \mhyph\ t'| & = & |t'| \\
    |\varphi\ t\ \mhyph\ t'\ \{t''\}| & = & |t''| &\ &
    |\chi\ T\ \mhyph\ t'| & = & |t'|
  \end{array}
  \]
  \caption{Erasure for annotated terms}
  \label{fig:eraser}
\end{figure}

@@

We have modified the rules for equality types $\{ t \simeq t' \}$ so that we require
nothing of $t$ and $t'$ except that the set $\textit{dom}(\Gamma)$ of variables
declared by $\Gamma$ includes their free variables $\textit{FV}(t\ t')$.  Further modifications
over the version of CDLE in~\cite{stump18} are:
\begin{itemize}
\item To prove $\{ t \simeq t\}$, one now writes $\beta\{t'\}$,
  with the critical idea that $|\beta\{t'\}|$ erases to $|t'|$.  We call this the \textbf{Kleene trick},
  because it goes back to Kleene's numeric
  realizability~\cite{Kle65_Classical-Extensions-of-Intuitionistic-Mathematics},
  which accepts any number $n$ as a realizer 
  of a true equation.  Here, we accept any closed term $t$ as a realizer
  of $\{ t \simeq t\}$.  This means that in Cedille, any such term -- even otherwise untypable
  terms, non-normalizing terms, etc. -- have type $\{ t \simeq t\}$ for any term
  $t$.
\item The \(\rho\) construct allows one to rewrite occurrences of \(t_1\) to
  \(t_2\) in the synthesized type of the second subexpression.
  The version presented here requires a ``guide'' \(@x.T\), where \(x\)
  indicates the occurrences of \(t_1\) to substitute.
  \([t_2/x]T\) must be a well-kinded type, but it is not required that
  \([t_1/x]T\) also be, only that \(T'\) and \([t_1/x]T\) are convertible modulo
  erasure --- it is not guaranteed that \(t_1\) and \(t_2\) will have the same
  type, or will even be typeable.
  In Cedille, the guide is optional and the construct may be used to rewrite a
  contextually given type; a heuristic is used to produce a resulting type that
  is well-kinded whose details are beyond the scope of this document.
\item We adopt a strong form of Nuprl's \textbf{direct computation rules}~\cite{constable+86}:
  If we have a term $t'$ of type $T$ and a proof $t$ that $\{ t' \simeq t''\}$, then we may conclude that
  $t''$ has type $T$ by writing the annotated term $\varphi\ t\ \mhyph\ t'\{t''\}$, which
  erases to $t''$.
\item Where the previous version of CDLE uses $\beta$-equivalence for (erased) terms, we here adopt $\beta\eta$-equivalence.  This
  allows us to observe in many cases that retyping functions are actually $\beta\eta$-equivalent to $\absu{\lambda}{x}{x}$.
  While $\beta\eta$-equivalence takes more work to incorporate into intrinsic 
  type theory~\cite{geuvers92},
  it raises no difficulties for our extrinsic one.
\item In this version, we add an explicit axiom $\delta$ saying that Church-encoded boolean \emph{true} is different from
  \emph{false}.
  In the implementation of Cedille, this is generalized to the rule where the
  proof \(t\) synthesizes an equation \(\{t_1 \simeq t_2\}\) in which \(|t_1|\)
  and \(|t_2|\) are separable using the \emph{B\"ohm-out algorithm}~\cite{BDPR79_Bohm-Algorithm}.

@@

Defining \verb|Top| to be $\{\absu{\lambda}{x}{x} \simeq
\absu{\lambda}{x}{x}\}$, we may assign \verb|Top| to any closed term \verb|t|,
including non-normalizing ones. In our annotated syntax, we write
\texttt{\(\beta\)\{t\}}. Even without this, the presence of $\delta$ in
combination with $\varphi$ allows us to type non-normalizing terms assuming an
erased argument $x$ of type $\{ \textit{tt} \simeq \textit{ff} \}$ for
Church-encoded booleans \textit{tt} and \textit{ff}. For example, $\delta\ \mhyph x$
has type \(\{\absu{\lambda}{x}{x} \simeq \absu{\lambda}{x}{x\ x}\}\), and with
$\varphi$ we can use this to type $\Omega$ by changing the typed term
\(\texttt{id}\ \cdot\texttt{True}\ \texttt{id}\), where \verb|True| is
\(\abs{\forall}{X}{\star}{X \to X}\). But failure of normalization does not
impinge on Theorem~\ref{thm:consis}. Extensional Martin-L\"of type theory (MLTT)
is also non-normalizing, for a very similar reason, but this fact does not contradict
its logical soundness~\cite{dybjer16}. In CDLE, the guarantees one gets about
the behavior of terms are expressed almost entirely in their types. If the types
are weak, then not much is guaranteed; but stronger types can guarantee
properties like normalization, as demonstrated by the following theorem:

@@

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpcheck T'}
          {\Gamma\vdash T\tpsynth \star & \Gamma\vdash t \tpcheck T & T \cong T'} 
\]
Using the IH for the first premise and the assumption that $\interp{T'}_{\sigma,\rho}$ is in $\mathcal{R}$ and hence defined,
we can apply the IH to the third premise to get $\interp{T}_{\sigma,\rho} = \interp{T'}_{\sigma,\rho}$.  Using this and the IH for second premise,
we get the desired conclusion, using also the definition of erasure.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpsynth T}
          {\Gamma\vdash T\tpsynth \star & \Gamma\vdash t \tpsynth T' & T \cong T'} 
\]
By the IH applied to the second premise, we have $[\sigma|t|]_{\cbe}\in\interp{T'}_{\sigma,\rho}\in\mathcal{R}$.
Using definedness of $\interp{T'}_{\sigma,\rho}$ and the IH applied to the first premise, we can apply the
IH to the third premise to get $\interp{T}_{\sigma,\rho} = \interp{T'}_{\sigma,\rho}$, from which the desired
conclusion follows by definition of erasure.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\{t''\} \Leftrightarrow T}
          {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}  
\]
By the IH for the first premise, $\sigma|t'| =_{\cbe} \sigma|t''|$.  By the IH for the second
premise, $[\sigma|t'|]_{\cbe}\in\interp{T}_{\sigma,\rho}$.  This suffices for the desired conclusion,
using also the definition of erasure ($|\varphi\ t\ \mhyph\ t'\{t''\}| = |t''|$).
